---
title: "Assignment 1: Reproducible Research"
author: "Benjamin Basow"
date: "October 1, 2017"
output: html_document
---

Setting the working directory and loading/viewing the data

```{r}
## Loading and Processing the Data 

activeData <- read.csv ("activity.csv", header = TRUE, sep = ",", stringsAsFactors = NA)

## Checking what the data looks like

summary(activeData)

str(activeData)

head(activeData)

## Modifying the date and interval variables

activeData$date <- as.Date(activeData$date, format = "%Y-%m-%d")
activeData$interval <- factor(activeData$interval)

## Checking the data set

head(activeData)

```

Assignment Prompt: What is the total number of steps taken per day?

```{r}
## Creating a data set that ignores NA values

NA_index <- is.na(as.character(activeData$steps))

cleanData <- activeData[!NA_index,]

# Checking clean data set

head(cleanData)

## Creating a variable that counts step per day

steps_per_day <- aggregate(steps ~ date, data = cleanData, sum)

## Checking new variable

head(steps_per_day)

## Explicitly adding names to the columns of new variable

colnames(steps_per_day) <- c("date", "steps")


## Making a histogram of the total number of steps taken each day

hist(as.numeric(steps_per_day$steps), breaks = 30, col = "blue", 
    
     xlab = "Number of Steps", 
     
     main= "Total number of steps taken per day")

## Calculating and reporting the mean and median of the total number of steps taken per day

mean(steps_per_day$steps)

median(steps_per_day$steps)

```

Assignment prompt: What is the average daily activity pattern?

```{r}

## Creating a variable that calculate the average number of steps taken per day 
## across all days

steps_all <- aggregate(cleanData$steps, by=list(interval=cleanData$interval), FUN=mean)

## Explicitly adding names to the columns of new variable

colnames(steps_all) <- c("interval", "average_steps")

## Plotting the average number of steps taken across all days

plot(as.integer(levels(steps_all$interval)), steps_all$average_steps, type="l",
     xlab = "Interval", ylab = "Average Number of Steps", main = "Average Daily Activity Pattern",  col ="slateblue")

## Creating variable for average max total steps

step_max<- max(steps_all$average_steps)

## checking step_max

step_max

# Calulating the 5-minute interval that contains the maximum number of steps

steps_max_interval<-steps_all[which.max(steps_all$average_steps),]$interval

## Checking the max 5 minute interval

steps_max_interval

## Assignment prompt: Which 5-minute interval, on average across all the days in the dataset, 
##                    contains the maximum number of steps?

## Response: interval 835 with the maximum amount of steps being 206.1698 or 206

```

Assignment prompt: Imputing missing values; Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}

sum(is.na(as.character(activeData$steps)))

sum(is.na(as.character(activeData$date)))

sum(is.na(as.character(activeData$interval)))

## only the steps column has missing data with 2304 missing observations

## Creating new data set with missing values replaced by the mean for the 5 minute interval

NA_index <- which(is.na(as.character(activeData$steps)))

data_final <- activeData

## Replace the NA values with the mean for the 5 minute interval data

data_final[NA_index, ]$steps<-unlist(lapply(NA_index, FUN=function(NA_index){
  
  steps_all[activeData[NA_index,]$interval==steps_all$interval,]$average_steps
  
}))

# Checking data_final

summary(data_final)

## Making a histogram of the total number of steps taken each day and Calculating and 
## reporting the mean and median total number of steps taken per day. 

#Creating a varable with the steps taken for each day

steps_each_day_final <- aggregate(steps ~ date, data = data_final, sum)

## Explicitly adding names to the columns of new variable

colnames(steps_each_day_final) <- c("date", "steps")

## Creating the Histogram with imputed values 

hist(as.numeric(steps_each_day_final$steps), breaks = 20, col = "tomato", 
     
     xlab = "Number of Steps", main= "Total number of steps taken each day")

## Calulating the mean and median

mean(steps_each_day_final$steps)

median(steps_each_day_final$steps)

## There is almost no difference in mean and median when we input values to replace the NA values.
## The mean is the same and the median is off by one step


```

Assignment prompt: Are there differences in activity patterns between weekdays and weekends?

```{r}
## Creating a day variable

data_final$day <- as.factor(weekdays(data_final$date))

## Creating a weekday variable

data_final$is_weekday <- ifelse(!(data_final$day %in% c("Saturday","Sunday")), TRUE, FALSE) 


## Calculating the average number of steps for weekdays

data_weekdays <- data_final[data_final$is_weekday,]

steps_per_interval_weekdays <- aggregate(data_weekdays$steps, by=list(interval=data_weekdays$interval), FUN=mean)


## Calculating the average number of steps for weekends

data_weekends <- data_final[!data_final$is_weekday,]

steps_per_interval_weekends <- aggregate(data_weekends$steps, by=list(interval=data_weekends$interval), FUN=mean)

## Explicitly adding column names

colnames(steps_per_interval_weekdays) <- c("interval", "average_steps")

colnames(steps_per_interval_weekends) <- c("interval", "average_steps")

## Adding new column for specific days

steps_per_interval_weekdays$day <- "Weekday"

steps_per_interval_weekends$day <- "Weekend"

## Creating final week variable

data_week_final <- rbind(steps_per_interval_weekends, steps_per_interval_weekdays)

## making the final day varable

data_week_final $day <- as.factor(data_week_final $day)

## Creating the plot

library(lattice)

xyplot(average_steps ~  interval | day, data = data_week_final , layout = c(1,2), type ="l", ylab="Number of Steps")

```




